---
title: "computational_flow_cytometry"
author: "Gabriel Rosenfeld"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

This tutorial will cover the basics of computational flow cytometry analysis. The goal is to introduce the fundamental aspects of such an analysis including reading/writing .fcs files, common data analysis steps, and visualizations. It is appropriate for those with a basic understanding of R and experience with flow cytometry data.

The course contains an R script from which the entire step by step analysis can be performed `computational_flow_cytometry.R` as well as this R markdown file in which the specific steps of the analysis can be observed while following along. This course expects R4.3.1+ is installed along with the complete set of R packages in the `installation.R` script.

## Set up

If the required R packages are not already installed on your machine, then uncomment and execute the R code below:

```{r}
# source("installation.R")
```

## Tutorial

Load the required libraries,

```{r}
library(flowCore)
library(ggcyto)
library(flowAI)
library(openCyto)
library(tidyverse)
library(magrittr)
```

Load Bcells tutorial dataset and save as object `fs` short for flowset,

```{r}
data("Bcells")
fs <- Bcells
```

The flowset object consists of individual flow frames. The `fs` object we are working with has 3 such flow frames in it.

```{r}
fs
```

Indexing into the flowset is similar to working with lists or vectors in R. Here we identify the first flow frame in this flowset and save it as an R object

```{r}
ff <- fs[[1]]
```

Lets briefly explore the data. Each flow frame has the same columns (corresponding to a channel on the flow cytometer that collected the data). Each channels has a range of values.

```{r}
flowCore::summary(fs)
```

This function also works on individual flow frames.

```{r}
flowCore::summary(ff)
```

Now that we know some of the parameters from the experiment, let's see how to index further. We'll also use piping in R which can help make complicated workflows easier.

```{r}
fs[, "Time"] %>%
flowCore::summary() # Here we pull data on only the time channel 

fs[2:3, "FSC-A"] %>%
flowCore::summary() # Here we pull data on only the FSC-A channel for samples 2 and 3
```

```{r}
# You can also find more information about the FCS file header
# This function can be used with individual flow frames for more details
keyword(ff)
```

### Reading and writing fcs files

Let's save our flowset. We can provide the directory name of bcells. And save in our current working directory.

```{r}
write.flowSet(x = fs, outdir = "bcells")
```

Now let's read in the data to a new flowset. We'll set the path as our previous saved folder. We'll look for files ending in .fcs. We'll load in the data in these files without any transformation.

```{r}
fs_read <- read.flowSet(path = "bcells", pattern = ".fcs", transformation = F, truncate_max_range = F)
```

Get high level information about the channels. We look at the parameters data from the first flow frame.

```{r}
channel_metadata <- fs[[1]]@parameters@data
View(channel_metadata)
```

You can also get just the color channels using `markernames` function. It only finds channels with the desc filled in (often what we care about but not always!).

```{r}
markers_of_interest <- markernames(fs)

markers_of_interest
```

### Data Quality Control

It's important to run a QC algorithm to remove any events that are low quality. For example, sudden changes in the flow cytometer performance due to clogging, laster fluctuations, detector spikes, etc.

```{r}
help("flow_auto_qc") # flowAI can be used for traditional flow cytometry data

flow_auto_qc(fcsfiles = fs, output = 2, folder_results = "resultsQC", html_report = F)

list.files("resultsQC") # QC fcs files and a .txt report
```

Let's read the .txt file describing the output of quality control output.

```{r}
read_delim(file = "resultsQC/QCmini.txt") %>% View()
```

Now we read the QC files.

```{r}
fs_qc <- read.flowSet(path = "resultsQC", pattern = ".fcs")

fs[[1]]
fs_qc[[1]] # the QC algorithm has added a marker of the low quality events

autoplot(fs_qc[[1]])
```

Let's filter out the poor quality events above 10000.

```{r}
qcGate <- rectangleGate(filterId = "highQuality","remove_from_all" = c(-Inf, 10000))
qc_filt <- flowCore::filter(x = fs_qc, qcGate)
fs_qc_filt <- flowCore::Subset(x = fs_qc, subset = qc_filt)

autoplot(fs_qc_filt[[1]]) # Only events under 10k remain in the remove_from_all channel
```

### Preprocessing

#### Compensation

For situations where the overlap of the fluorochromes occurs, it is important to run compensation. If available, you may find the compensation matrix directly in the file using the keywords.

```{r}
comp <- keyword(fs_qc_filt[[1]])[["SPILL"]]
comp # The matrix provides the information on how much signal to subtract out of one channel from another

keyword(fs_qc_filt[[1]])[["APPLY COMPENSATION"]] # Compensation is already performed so we don't need to
```

The `fsApply` function can apply a flow cytometry function to each flow frame within a flow set. Let's use this function to decompensate the data and we'll compare the effect with the compensated data.

```{r}
fs_qc_filt_decomp <- fsApply(fs_qc_filt, FUN = function(ff) decompensate(ff, comp))
```

We can use the ggcyto package to do ggplot2 type visualizations! Here we compare a couple of channels with overlap from the compensation matrix with biological significance.

```{r}
plot_comp <- ggcyto(fs_qc_filt[[1]], aes("Pacific", "655-A")) + geom_hex(bins = 512) + scale_x_flowjo_biexp() + scale_y_flowjo_biexp() + ggtitle("Compensated")
plot_uncomp <- ggcyto(fs_qc_filt_decomp[[1]], aes("Pacific", "655-A")) + geom_hex(bins = 512) + scale_x_flowjo_biexp() + scale_y_flowjo_biexp() + ggtitle("Uncompensated")
```

```{r}
plot_uncomp + ggtitle("Uncompensated")
```

```{r}
plot_comp + ggtitle("Compensated")
```

#### Additional Preprocessing

Let's also remove any margin events (not always needed and depends upon experiment). We remove from scatter channels.

```{r}
bf <- boundaryFilter(x = c("FSC-A", "SSC-A", "FSC-H", names(markers_of_interest)), filterId = "boundaryFilt")
fs_qc_filt_margins <- flowCore::Subset(x = fs_qc_filt, subset = bf)
```

#### Data visualization and Transformation

On the untransformed data, it is hard to see differences across the color channels without transformation. Also note that the remove_from_all QC channel is limited to below 10000.

```{r}
autoplot(fs_qc_filt_margins[[1]])
```

Logicle Transformation is one approach to allowing better visualization of the color channels.

```{r}
trans <- logicle_trans()
trans <- transformList(from = names(markers_of_interest), logicleTransform())

fs_qc_filt_margins_trans <- transform(fs_qc_filt_margins, trans)
```

Now the plots for the color channels are transformed for visualization

```{r}
autoplot(fs_qc_filt_margins_trans[[1]])
```

### Gating

Let's use `openCyto` to perform some common gating steps in a more automated fashion.

```{r}
gs <- GatingSet(fs_qc_filt_margins_trans)
```

Let's examine our cells by FSC and SSC.

```{r}
ggcyto(data = gs, aes(`FSC-A`, `SSC-A`), subset = "root") + geom_hex(bins = 512)
```

We can gate out debris by a removal of events around \~1e05 SSC.

```{r}
gs_add_gating_method(gs = gs, alias = "nonDebris", dims = "FSC-A", pop = "+",
                     parent = "root", gating_method = "mindensity", gating_args = "peaks = c(0,1.25e05)")

ggcyto(gs, aes("FSC-A", "SSC-A"), subset = "root") + geom_gate("nonDebris") + geom_hex(bins = 512) + geom_stats()
```

Now we add a lymphocytes gate around the events with debris removed.

```{r}
gs_add_gating_method(gs, alias = "lymphocytes", pop = "+", parent = "nonDebris", dims = "FSC-A,SSC-A",
                     gating_method = "flowClust.2d", gating_args = "K=2, target=c(1e05,5e04),quantile=0.95"
                    )

ggcyto(gs, aes("FSC-A", "SSC-A"), subset = "nonDebris") + geom_gate("lymphocytes") + geom_hex(bins = 512) + geom_stats()
```

Let's add a single gate on the lymphocytes population.

```{r}
gs_add_gating_method(gs, alias = "singlets", pop = "+", parent = "lymphocytes", dims = "FSC-A,FSC-H",
                     gating_method = "singletGate")

ggcyto(gs, aes("FSC-A", "FSC-H"), subset = "lymphocytes") + geom_gate("singlets") + geom_hex(bins = 512) + geom_stats()
```

Finally, we gate CD19+ events on the singlets.

```{r}
ggcyto(gs, aes(x="CD19"), subset = "singlets") + geom_density()

gs_add_gating_method(gs, alias = "bcells", pop = "+", parent = "singlets", dims = "CD19",
                     gating_method = "mindensity")

ggcyto(gs, aes("CD19"), subset = "singlets") + geom_gate("bcells") + geom_density() + geom_stats()
```

Now let's do some statistical analysis on the means and median in each gated population.

```{r}
gs_stats <- gs_pop_get_count_fast(x = gs, type = "count")
gs_stats <- left_join(gs_stats, gs_pop_get_count_fast(x = gs, statistic = "freq"))
gs_stats <- dplyr::mutate(gs_stats, freq_of_parent = Count/ParentCount)

pop.median <- function(fr){
  chnls <- colnames(fr)
  res <- matrixStats::colMedians(exprs(fr))
  names(res) <- chnls
  res
}

pop.mean <- function(fr){
  chnls <- colnames(fr)
  res <- matrixStats::colMeans2(exprs(fr))
  names(res) <- chnls
  res
}

gs_stats_mean <-  left_join(gs_stats, gs_pop_get_stats(x = gs, type = pop.mean) %>% dplyr::rename("name" = "sample", "Population" = "pop"))
gs_stats_median <-  left_join(gs_stats, gs_pop_get_stats(x = gs, type = pop.median) %>% dplyr::rename("name" = "sample", "Population" = "pop"))
```
